---
title: "How to Analyze Probabilistic Monitoring Data (multiple parameters) for Integrated Report Chapter"
author: "Emma Jones"
date: "March, 2020"
output: html_document
---

```{r startup, echo=FALSE, warning=FALSE}
# R version 3.6.1 "Action of the Toes"
suppressPackageStartupMessages(library(tidyverse))#1.3.0
suppressPackageStartupMessages(library(sf))#0.8-0
suppressPackageStartupMessages(library(spsurvey))#4.1.1
suppressPackageStartupMessages(library(DT))#0.11
suppressPackageStartupMessages(library(purrr))#0.3.3
```

This document walks you through the steps of calculating design weights, running analyses, and writing the Integrated Report chapter for the Probabilistic Monitoring Program due to EPA every two years. This document overviews the 2020 IR ProbMon chapter.




### Data Organization

#### Landcover Organization

Landcover metrics were completely rerun between the 2018 IR and 2020IR to update watershed delineation methods to the StreamStats autodelineate watersheds and to streamline/improve certain metrics like landcover (updated years NLCD applied to) and road (pull annual TIGERroad data). All 2001-2018 wadeable watersheds were visually inspected by Emma Jones. 

Below is the combination of all landcover data prior to smashing with biology/chemistry data.

```{r landcover smash}
files <- c(paste0('C:/HardDriveBackup/R/GitHub/LandcoverAnalysis/Results/ProbWadeable2001_2012/',
                  2001:2012,'/Result10.csv'),
           paste0('C:/HardDriveBackup/R/GitHub/LandcoverAnalysis/Results/ProbWadeable2013_2018/', 
                  2013:2018,'/Result10.csv')) 

landcover <- files %>%
  map(read_csv) %>%    # read in all the files individually
  reduce(bind_rows) %>% # combine
  mutate(Year = YearSampled) # joining column


```


#### Wadeable Bio/Chem Results

First smash two more years of data to existing dataset.

(add next two years was my initial smash, then more QA back and forth with Jason led to the following chunk named final data smash)

```{r add next two years, eval= F}
existingData <- read_csv('originalData/Wadeable_ProbMon_2001-2016_EVJ.csv') %>%
  mutate(IR2020 = NA) %>% # add new column
  # manual column reorganization &drop old landcover info bc we will replace with autodelineated
  dplyr::select(DataSource , StationID , Year , StationID_Trend , LongitudeDD , LatitudeDD , stratum , designweight , weightcategory , station , state , status , comment , set , Basin , SubBasin , BayShed , BayPanel , EcoRegion , BioRegion , Panel , BioPanel , Order , BasinSize , StreamSizeCat , StreamSizeCatPhase , AREA_SQ_MILES , IR2008 , IR2010 , IR2012 , IR2014 , IR2016 , IR2018 , IR2020 , designweighttrend , designweightoriginal , filwgt_trend , filwgt_orgil , DO , pH , SpCond , TN , TP , TDS , NH4 , NO3 , TKN , `Ortho-P` , Turb , TSS , Na , K , Cl , Sf , `70331VFine` , SSCCOARSE , SSCFINE , SSCTOTAL , LRBS , Slope , FN_PCT , SA_PCT , SA_FN_PCT , LSUB_DMM , BL_CB_GR_Embed_PCT , Embed_PCT , TotHab , TotTaxa , EPTTax , VEphem , VPTHydropsychidae , VScrap , VChiro , V2Dom , HBI , VHapto , EPTInd , VSCIVCPMI , MetalCCU , ARSENICppm , BERYLLIUMppm , CADMIUMppm , CHROMIUMppm , COPPERppm , LEADppm , MANGppm , NICKELppm , SILVERppm , ZINCppm , ANTIMONYppm , ALUMINUMppm , SELENIUMppm , IRONppm , MERCURYppm , THALLIUMppm , CALCIUM , MAGNESIUM , ARSENIC , BARIUM , BERYLLIUM , CADMIUM , CHROMIUM , COPPER , IRON , LEAD , MANGANESE , THALLIUM , NICKEL , SILVER , ZINC , ANTIMONY , ALUMINUM , SELENIUM , HARDNESS , MERCURY , `Hg-C`)

# watershed results this cycle
newData <- readxl::read_excel('originalData/Wadeable_ProbMon_20017-2018.xlsx',
                              sheet = 'All20172018Data') %>%
  mutate(Year = as.numeric(Year)) %>%
   dplyr::select(DataSource , StationID , Year , StationID_Trend , LongitudeDD , LatitudeDD , stratum , designweight , weightcategory , station , state , status , comment , set , Basin , SubBasin , BayShed , BayPanel , EcoRegion , BioRegion , Panel , BioPanel , Order , BasinSize , StreamSizeCat , StreamSizeCatPhase , AREA_SQ_MILES , IR2008 , IR2010 , IR2012 , IR2014 , IR2016 , IR2018 , IR2020 , designweighttrend , designweightoriginal , filwgt_trend , filwgt_orgil , DO , pH , SpCond , TN , TP , TDS , NH4 , NO3 , TKN , `Ortho-P` , Turb , TSS , Na , K , Cl , Sf , `70331VFine` , SSCCOARSE , SSCFINE , SSCTOTAL , LRBS , Slope , FN_PCT , SA_PCT , SA_FN_PCT , LSUB_DMM , BL_CB_GR_Embed_PCT , Embed_PCT , TotHab , TotTaxa , EPTTax , VEphem , VPTHydropsychidae , VScrap , VChiro , V2Dom , HBI , VHapto , EPTInd , VSCIVCPMI , MetalCCU , ARSENICppm , BERYLLIUMppm , CADMIUMppm , CHROMIUMppm , COPPERppm , LEADppm , MANGppm , NICKELppm , SILVERppm , ZINCppm , ANTIMONYppm , ALUMINUMppm , SELENIUMppm , IRONppm , MERCURYppm , THALLIUMppm , CALCIUM , MAGNESIUM , ARSENIC , BARIUM , BERYLLIUM , CADMIUM , CHROMIUM , COPPER , IRON , LEAD , MANGANESE , THALLIUM , NICKEL , SILVER , ZINC , ANTIMONY , ALUMINUM , SELENIUM , HARDNESS , MERCURY , `Hg-C`)

# Since added IR2020 column to existingData have to reorganize column order before smash
names(existingData) == names(newData) 
#names(existingData)[31:40]
#names(newData)[31:40]
#names(existingData) <- names(newData)  

existingData <- mutate(existingData, MERCURY = as.numeric(as.character(MERCURY)))

allData <- bind_rows(existingData, newData)
rm(existingData); rm(newData)
```


This is Jason's QAed version of what I did in the previous step.

```{r final data smash}
allData <- readxl::read_excel('processedData/Wadeable_ProbMon_2001-2018_Final_Final.xlsx') %>%
  dplyr::select(-c(YearSampled:STXRD))# drop old landcover
```


Need to join landcover data to 2001-2018 data


```{r landcover smash}
allData <- left_join(allData, landcover, by = c('StationID', 'Year')) %>%
  # fix that only 2017-2018 have IR2020 designation
  mutate(IR2020 = case_when(Year %in% c(2013:2018) ~ 2020, TRUE ~ 1))

#Save for now
#write.csv(allData, 'processedData/Wadeable_ProbMon_2001-2018_Final_Final.csv', row.names = F)
rm(landcover); rm(files); rm(allData)
```


#### Design Status

Starting with an organized 'design status' database that has all Probmon stations (sampled or not) organized with stream order information. This data is important to note all sites that were evaluated and whether or not they should be included in the design weight process. Environmental, chemical, landcover, etc. data is organized in a separate spreadsheet (brought in later) that is joined to the weight category results. 

Bring in appropriate datasets. Always start with VSCI (VSCI/VCPMI) first because this parameter will have the best information on whether or not to include the station for design weight calculations. 
```{r bringInTSData}

#### Read in master field data sampled at all sites
# Select the parameters we want to run
# rename a few parameters to work better in R
allData <- read_csv('processedData/Wadeable_ProbMon_2001-2018_Final_Final.csv')

library(magrittr)
metals <- names(select(allData,MetalCCU:MERCURY))
allData[,metals] %<>% lapply(function(x) as.numeric(as.character(x)))

```

### Metals Analysis

MetalsCCU calculation function

```{r metalsCCUfunction}
# Capitalize first letter in a word easily
capwords <- function(s, strict = FALSE) {
  cap <- function(s) paste(toupper(substring(s, 1, 1)),
                           {s <- substring(s, 2); if(strict) tolower(s) else s},
                           sep = "", collapse = " " )
  sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
}

# Metals CCU Calculation
metalsCCUcalc <- function(Hardness,Aluminum,Arsenic,Cadmium,Chromium,Copper,Lead,Nickel,Selenium,Zinc){
  criteriaHardness <- ifelse(Hardness<25,25,ifelse(Hardness>400,400,Hardness))
  AluminumEPAChronic <- Aluminum/150
  ArsenicChronic <- Arsenic/150
  CadmiumChronic <- Cadmium/(exp(0.7852*(log(criteriaHardness))-3.49))
  ChromiumChronic <- Chromium/((exp(0.819*(log(criteriaHardness))+0.6848))*0.86)
  CopperChronic <- Copper/((exp(0.8545*(log(criteriaHardness))-1.702))*0.96)
  LeadChronic <- Lead/((exp(1.273*(log(criteriaHardness))-3.259)))
  NickelChronic <- Nickel/((exp(0.846*(log(criteriaHardness))-0.884))*0.997)
  SeleniumChronic <- Selenium/5
  ZincChronic <- Zinc/((exp(0.8473*(log(criteriaHardness))+0.884))*0.986)
  return(sum(AluminumEPAChronic,ArsenicChronic,CadmiumChronic,ChromiumChronic,CopperChronic,
             LeadChronic,NickelChronic,SeleniumChronic,ZincChronic))
}

# Metals CCU Calculation for dataframes
# changed slightly from Benthic Stressor Tool
metalsCCUcalcDF <- function(df){
  siteID <- df$StationID_Trend # changed
  Year <- df$Year #changed
  Hardness <- df$Hardness
  Aluminum <- df$Aluminum
  Arsenic <- df$Arsenic
  Cadmium <- df$Cadmium
  Chromium <- df$Chromium
  Copper <- df$Copper
  Lead <- df$Lead
  Nickel <- df$Nickel 
  Selenium <- df$Selenium
  Zinc <- df$Zinc
  met <- data.frame(StationID_Trend = NA, Year=NA,MetalCCU=NA) #changed
  for(i in 1:nrow(df)){
    met[i,] <- cbind(as.character(siteID[i]),as.character(Year[i]),
                     format(metalsCCUcalc(df$Hardness[i],df$Aluminum[i],df$Arsenic[i],
                                          df$Cadmium[i],df$Chromium[i],df$Copper[i],
                                          df$Lead[i],df$Nickel[i],df$Selenium[i],
                                          df$Zinc[i]),digits=4))
  }
  return(met)
}
```

```{r metalsCCUcalculation}
metalsCCUresults <- select(allData,StationID_Trend, Year, HARDNESS, ALUMINUM, ARSENIC, CADMIUM, 
                           CHROMIUM, COPPER, LEAD, NICKEL, SELENIUM, ZINC)
names(metalsCCUresults)[3:length(metalsCCUresults)] <- capwords(tolower(names(metalsCCUresults)[3:length(metalsCCUresults)]))
metalsCCUresults <- metalsCCUcalcDF(metalsCCUresults)
metalsCCUresults$Year <- as.numeric(as.character(metalsCCUresults$Year))

# quick double check to make sure all the same as previous 
View(left_join(dplyr::select(allData, StationID_Trend, Year, MetalCCU),
               metalsCCUresults, by = 'StationID_Trend') %>%
       mutate(diff = as.numeric(MetalCCU.y) - MetalCCU.x) %>%
       arrange(diff))

allDataMCCU <- select(allData,-c(MetalCCU)) %>%
  left_join(metalsCCUresults,by=c('StationID_Trend','Year'))%>%
  select(DataSource,StationID,Year,StationID_Trend,LongitudeDD,LatitudeDD,stratum,designweight,
         weightcategory,station,state,status,comment,set,Basin,SubBasin,BayShed,BayPanel,
         EcoRegion,BioRegion,Panel,BioPanel,Order,BasinSize,StreamSizeCat,StreamSizeCatPhase,
         AREA_SQ_MILES,
         IR2008,IR2010,IR2012,IR2014,IR2016,IR2018,IR2020, 
         DO,pH,SpCond,TN,TP,TDS,NH4,
         NO3,TKN,`Ortho-P`,Turb,TSS,Na,K,Cl,
         Sf,`70331VFine`,SSCCOARSE,SSCFINE,SSCTOTAL,LRBS,
         Slope,FN_PCT,SA_PCT,SA_FN_PCT,LSUB_DMM,BL_CB_GR_Embed_PCT,Embed_PCT,
         TotHab,TotTaxa,EPTTax,VEphem,VPTHydropsychidae,
         VScrap,VChiro,V2Dom,HBI,VHapto,EPTInd,VSCIVCPMI,MetalCCU,everything())

#write.csv(allDataMCCU,'processedData/Wadeable_ProbMon_2001-2018.csv', row.names = F) # overwrite previous
# then manually save to .xlsx for jason
``` 


Run only parameters we want at present and get ready for CDF

```{r}
allData <- read_csv('processedData/Wadeable_ProbMon_2001-2018.csv')
surveyData <- select(allData, StationID_Trend,Year,DO:MetalCCU,CALCIUM:MERCURY,wshdImpPCT) %>% 
  dplyr::rename(siteID=StationID_Trend,Ortho_P= "Ortho-P",
                X70331VFine=`70331VFine`)

```



Now time to find any exceedances of metals criteria for acute or chronic.

```{r}
metalsData <- select(surveyData, siteID,Year, CALCIUM:MERCURY)
names(metalsData)[3:22] <- capwords(tolower(names(metalsData)[3:22]))

metalsCriteriaAssessment <- function(df){
  df1 <- gather(df, Metal, Measure, Calcium:Hardness) 
  
  criteriaHardness <- ifelse(df$Hardness<25,25,ifelse(df$Hardness>400,400,df$Hardness))
  x <- data.frame(Calcium_Chronic=NA,Calcium_Acute=NA,Calcium_PWS=NA,
                  Magnesium_Chronic=NA, Magnesium_Acute=NA, Magnesium_PWS=NA,
                  Arsenic_Chronic=150,Arsenic_Acute=340,Arsenic_PWS=10,
                  Barium_Chronic=NA,Barium_Acute=NA,Barium_PWS=2000,
                  Beryllium_Chronic=NA,Beryllium_Acute=NA,Beryllium_PWS=NA,
                  Cadmium_Chronic=format((exp(0.7852*(log(criteriaHardness))-3.49)),digits=3), 
                  Cadmium_Acute= format((exp(1.128*(log(criteriaHardness))-3.828)),digits=3), Cadmium_PWS=5,
                  Chromium_Chronic=format(((exp(0.819*(log(criteriaHardness))+0.6848))*0.86),digits=3),
                  Chromium_Acute=format(((exp(0.819*(log(criteriaHardness))+3.7256))*0.316),digits=3), Chromium_PWS=100,
                  Copper_Chronic=format(((exp(0.8545*(log(criteriaHardness))-1.702))*0.96),digits=3),
                  Copper_Acute=format(((exp(0.9422*(log(criteriaHardness))-1.70))*0.96),digits=3), Copper_PWS=1300,
                  Iron_Chronic=NA, Iron_Acute=NA, Iron_PWS= 300,
                  Lead_Chronic=format(((exp(1.273*(log(criteriaHardness))-3.259))),digits=3),
                  Lead_Acute=format(((exp(1.273*(log(criteriaHardness))-1.084))),digits=3), Lead_PWS=15,
                  Manganese_Chronic=NA, Manganese_Acute=NA, Manganese_PWS=50,
                  Thallium_Chronic=NA, Thallium_Acute=NA, Thallium_PWS=0.24,
                  Nickel_Chronic=format(((exp(0.846*(log(criteriaHardness))-0.884))*0.997),digits=3),
                  Nickel_Acute=format(((exp(0.846*(log(criteriaHardness))+1.312))*0.998),digits=3),Nickel_PWS=610,
                  Silver_Chronic=NA, Silver_Acute=format(((exp(1.72*(log(criteriaHardness))-6.52))*0.85),digits=3), Silver_PWS=NA,
                  Zinc_Chronic=format(((exp(0.8473*(log(criteriaHardness))+0.884))*0.986),digits=3),
                  Zinc_Acute=format(((exp(0.8473*(log(criteriaHardness))+0.884))*0.978),digits=3), Zinc_PWS=7400,
                  Antimony_Chronic=NA, Antimony_Acute=NA, Antimony_PWS=5.6,
                  Aluminum_Chronic=NA,Aluminum_Acute=NA,Aluminum_PWS=NA,
                  Selenium_Chronic=5, Selenium_Acute= 20, Selenium_PWS=170,
                  Hardness_Chronic=NA,Hardness_Acute=NA,Hardness_PWS=NA)
  x1 <- as.data.frame(t(x)) %>%
    rownames_to_column('original') %>%
    rowwise()%>%  
    mutate(Metal=gsub("_.*","\\1",original),Criteria=gsub(".*_","\\1",original)) %>%
                            group_by(Criteria) %>%
                            mutate(id = row_number()) %>%
                            select(-original) %>%
                            spread(Criteria, V1) %>%
                            select(-id) %>%
    full_join(df1, by= 'Metal') 
  x1[,c(2:4)] %<>% lapply(function(x) as.numeric(as.character(x)))
  x2 <- mutate(x1, Acute_Exceed=ifelse(Measure>Acute,1,0),
           Chronic_Exceed=ifelse(Measure>Chronic,1,0),
           PWS_Exceed=ifelse(Measure>PWS,1,0)) %>%
      select(siteID,Year,Metal,Measure,Acute,Chronic,PWS,Acute_Exceed,Chronic_Exceed,PWS_Exceed)
  return(x2)
}

metalsAssessment <- data.frame(siteID=NA,Year=NA,Metal=NA,Measure=NA,Acute=NA,Chronic=NA,
                               PWS=NA,Acute_Exceed=NA,Chronic_Exceed=NA,PWS_Exceed=NA)
for(i in 1:nrow(metalsData)){
  metalsAssessment <- rbind(metalsAssessment,suppressWarnings(metalsCriteriaAssessment(metalsData[i,])))
}
metalsAssessment <- arrange(metalsAssessment,siteID, Year)

# Any Acute Exceedances
filter(metalsAssessment, Acute_Exceed==1)

# Any Chronic Exceedances
filter(metalsAssessment, Chronic_Exceed==1)

# IR window Acute Exceedances
filter(metalsAssessment, Year >= 2013 & Year <= 2018 & Acute_Exceed==1)

# IR Window Chronic Exceedances
filter(metalsAssessment, Year >= 2013 & Year <= 2018 & Chronic_Exceed==1)


```


### CDF analyses


And now bring in design status data (for VSCI/VCPMI). It is best to use this as the default design status dataset (and adjust TS to OT if not sampled) because it is the most complete version of the data.
```{r bringInDesignStatusData}
designStatus <- readxl::read_excel('processedData/biology20012018final.xlsx',sheet='biology2020') %>%
  mutate(
    IR2008 = replace(IR2008, IR2008==1, NA),
    IR2010 = replace(IR2010, IR2010==1, NA),
    IR2012 = replace(IR2012, IR2012==1, NA),
    IR2014 = replace(IR2014, IR2014==1, NA),
    IR2016 = replace(IR2016, IR2016==1, NA),
    IR2018 = replace(IR2018, IR2018==1, NA),
    IR2020 = replace(IR2020, IR2020==1, NA),
    Panel1 = ifelse(Panel == "Phase1", 1, NA),
    Panel2 = ifelse(Panel == "Phase2", 1, NA),
    BioPanel1 = ifelse(BioPanel == "Phase1", 1, NA),
    BioPanel2 = ifelse(BioPanel == "Phase2", 1, NA),
    BioPanel3 = ifelse(BioPanel == "Phase3", 1, NA),
    BioPanel4 = ifelse(BioPanel == "Phase4", 1, NA)) %>% # housekeeping: recode biophase and change 1's to NA's to indicate it wasnt sampled in that window, break up Panel and BioPanel windows to separate columns for weight adjustment purposes
  dplyr::rename(siteID=sampleID ) # start playing nicely with spsurvey)


```



These are the data windows we are looking at for 2018 IR:

* Full sample window (2001-2018)
* 2020 IR (2013-2018)
* 2018 IR (2011-2016)
* 2016 IR (2009-2014)
* 2014 IR (2007-2012)
* 2012 IR (2005-2010)
* 2010 IR (2003-2008)
* 2008 IR (2001-2007)
* Phases (Phase 1= 2001-2008; Phase 2= 2009-2018) Kept Phase 1 same as 2018IR to balance n samples in window
* BioPhases (Phase 1 = 2001-2004; Phase 2 = 2005-2008; Phase 3 = 2009-2013; Phase 4 = 2014-2018;)



#### Functions for Weight Adjustments and run all CDF data by subpopulations

The following functions run the CDF analyses for all parameters chosen from the surveyData dataframe (you can modify the selection to increase/decrease parameters run). The subpopEstimate() function runs each subpopulation and nests inside the listOfResults(), which outputs a list of all subpopulation results. The allCDFresults() **adjusts all weights according to sample window** and then calls the listOfResults() to run all the CDF, percentile, and population estimates for each subpopulation. See below for how to run each of these functions and how to view outputs.

```{r allFunctions}
## Subpopulation estimates by parameter
subpopEstimate <- function(finalData,parameter, subpopulationCategory, subpopulation, altName, specialWeight){
  # Build in catch in case whole population needed
  if(is.na(subpopulationCategory) & subpopulation == "Virginia"){
    subpopData <- finalData
    sites.ext <- select(subpopData, siteID) %>% mutate(Use = TRUE)
    subpop.ext <- select(subpopData,siteID) %>% mutate(Region = 'Virginia')
  }else{
    subpopData <- finalData[ finalData[[subpopulationCategory]] == subpopulation, ] 
    if(nrow(subpopData) == nrow(finalData)){ # special catch for IR windows not filtering correctly
      subpopData <- finalData[ !is.na(finalData[[subpopulationCategory]] ), ] 
    }}
  
  # If no data in subpopulation, keep moving
  if(nrow(subpopData) !=0){
    sites.ext <- select(subpopData, siteID) %>% mutate(Use = TRUE)
    # Special Cases to match existing terminology for each Subpopulation
    if(is.na(altName)){
      subpop.ext <- select(subpopData,siteID) %>% mutate(Region = subpopulation)
    }else{
      subpop.ext <- select(subpopData,siteID) %>% mutate(Region = altName)
    }
    
   
    
  # Choose correct final weight and filter to stratum = 1
    if(specialWeight==FALSE){
      finalweight <- subpopData$finalweight_all
    }else{
      finalweight <- as.numeric(as.matrix(subpopData[,specialWeight]))
    }

  
  design.ext <- mutate(subpopData,siteID = siteID, stratum = "1", wgt = finalweight, xcoord = xmarinus, ycoord = ymarinus) %>%
    select(siteID, stratum, wgt, xcoord, ycoord)
  
  data.cont.ext <- select(finalData, siteID, parameter)
  
  subpopStatus <- cont.analysis(sites = data.frame(sites.ext), subpop = data.frame(subpop.ext), design = data.frame(design.ext), 
                                data.cont = data.frame(data.cont.ext),
                                pctval=c(1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, 90, 95, 99), 
                                conf=95, vartype="Local")
  return(subpopStatus)}else{return(list())}
}

# Now build it all into a single function that returns a list of lists (one list object per estimate, 
# each estimate is a list of of 4 dataframes where the important ones ($CDF and $Pct) can easily be 
# reached with list calls noted at bottom of page)
listOfResults <- function(popstatus.est, finalData, parameterName){
  list(
    popstatus.est = popstatus.est,
    dataAnalyzed = finalData,
    # All Virginia
    estimateVirginia = subpopEstimate(finalData,parameterName,NA, 'Virginia',NA, specialWeight=FALSE),
    # By Basin
    estimateRoanoke = subpopEstimate(finalData,parameterName, 'Basin', 'Roanoke','Roanoke Basin', specialWeight=FALSE),
    estimateJames = subpopEstimate(finalData,parameterName, 'Basin', 'James','James Basin', specialWeight=FALSE) ,
    estimatePotomacShenandoah = subpopEstimate(finalData,parameterName, 'Basin', 'Potomac-Shenandoah',NA, specialWeight=FALSE) ,
    estimateRappahannockYork = subpopEstimate(finalData,parameterName, 'Basin', 'Rappahannock-York',NA, specialWeight=FALSE),
    estimateNew = subpopEstimate(finalData,parameterName, 'Basin', 'New',NA, specialWeight=FALSE) ,
    estimateChowan = subpopEstimate(finalData,parameterName, 'Basin', 'Chowan',NA, specialWeight=FALSE), 
    estimateTennessee = subpopEstimate(finalData,parameterName, 'Basin', 'Tennessee',NA, specialWeight=FALSE), 
    estimateHolston = subpopEstimate(finalData,parameterName, 'SubBasin', 'Holston',NA, specialWeight=FALSE) ,
    estimateBigSandy = subpopEstimate(finalData,parameterName, 'SubBasin', 'Big Sandy',NA, specialWeight=FALSE) ,
    estimateClinchPowell = subpopEstimate(finalData,parameterName, 'SubBasin', 'Clinch-Powell',NA, specialWeight=FALSE) ,
    estimatePotomac = subpopEstimate(finalData,parameterName, 'SubBasin', 'Potomac',NA, specialWeight=FALSE) ,
    estimateShenandoah = subpopEstimate(finalData,parameterName, 'SubBasin', 'Shenandoah',NA, specialWeight=FALSE) ,
    estimateRappahannock = subpopEstimate(finalData,parameterName, 'SubBasin', 'Rappahannock',NA, specialWeight=FALSE) ,
    estimateYork = subpopEstimate(finalData,parameterName, 'SubBasin', 'York',NA, specialWeight=FALSE) ,
    # By Ecoregion
    estimatePiedmont = subpopEstimate(finalData,parameterName, 'EcoRegion', 'Piedmont',NA, specialWeight=FALSE) ,
    estimateNorthernPiedmont = subpopEstimate(finalData,parameterName, 'EcoRegion', 'Northern Piedmont',NA, specialWeight=FALSE) ,
    estimateCARV = subpopEstimate(finalData,parameterName, 'EcoRegion', 'Central Appalachian Ridges and Valleys',NA, specialWeight=FALSE) ,
    estimateSEplains = subpopEstimate(finalData,parameterName, 'EcoRegion', 'Southeastern Plains',NA, specialWeight=FALSE) ,
    estimateBRM = subpopEstimate(finalData,parameterName, 'EcoRegion', 'Blue Ridge Mountains',NA, specialWeight=FALSE) ,
    #estimateMACP = subpopEstimate(finalData,parameterName, 'EcoRegion', 'Middle Atlantic Coastal Plain',NA, specialWeight=FALSE) , # need 25-30 samples to run this analysis
    estimateCentralApps = subpopEstimate(finalData,parameterName, 'EcoRegion', 'Central Appalachians',NA, specialWeight=FALSE) ,
    # By Bioregion
    estimateMountainBioregion = subpopEstimate(finalData,parameterName, 'BioRegion', 'Mountain','Mountain Bioregion', specialWeight=FALSE),
    estimatePiedmontBioregion = subpopEstimate(finalData,parameterName, 'BioRegion', 'Piedmont','Piedmont Bioregion', specialWeight=FALSE) ,
    estimateCoastBioregion = subpopEstimate(finalData,parameterName, 'BioRegion', 'Coast', 'Coast Bioregion', specialWeight=FALSE),
    # By Order
    estimateFirstOrder = subpopEstimate(finalData,parameterName, 'Order', '1','First Order', specialWeight=FALSE) ,
    estimateSecondOrder = subpopEstimate(finalData,parameterName, 'Order', '2','Second Order', specialWeight=FALSE) ,
    estimateThirdOrder = subpopEstimate(finalData,parameterName, 'Order', '3','Third Order', specialWeight=FALSE) ,
    estimateFourthOrder = subpopEstimate(finalData,parameterName, 'Order', '4','Fourth Order', specialWeight=FALSE) ,
    estimateFifthOrder = subpopEstimate(finalData,parameterName, 'Order', '5','Fifth Order', specialWeight=FALSE) ,
    # By Basin Size
    estimateBasin1 = subpopEstimate(finalData,parameterName, 'BasinSize', '1','<1 square mile', specialWeight=FALSE) ,
    estimateBasin2 = subpopEstimate(finalData,parameterName, 'BasinSize', '2','1 to 10 square mile', specialWeight=FALSE) ,
    estimateBasin3 = subpopEstimate(finalData,parameterName, 'BasinSize', '3','10 to 200 square mile', specialWeight=FALSE) ,
    estimateBasin4 = subpopEstimate(finalData,parameterName, 'BasinSize', '4','>200 square mile', specialWeight=FALSE) ,
    # By Year
    estimate2001 = subpopEstimate(finalData,parameterName, 'Year', '2001','Year 2001', specialWeight='finalweight_Year') ,
    estimate2002 = subpopEstimate(finalData,parameterName, 'Year', '2002','Year 2002', specialWeight='finalweight_Year'), 
    estimate2003 = subpopEstimate(finalData,parameterName, 'Year', '2003','Year 2003', specialWeight='finalweight_Year') ,
    estimate2004 = subpopEstimate(finalData,parameterName, 'Year', '2004','Year 2004', specialWeight='finalweight_Year') ,
    estimate2005 = subpopEstimate(finalData,parameterName, 'Year', '2005','Year 2005', specialWeight='finalweight_Year') ,
    estimate2006 = subpopEstimate(finalData,parameterName, 'Year', '2006','Year 2006', specialWeight='finalweight_Year') ,
    estimate2007 = subpopEstimate(finalData,parameterName, 'Year', '2007','Year 2007', specialWeight='finalweight_Year') ,
    estimate2008 = subpopEstimate(finalData,parameterName, 'Year', '2008','Year 2008', specialWeight='finalweight_Year') ,
    estimate2009 = subpopEstimate(finalData,parameterName, 'Year', '2009','Year 2009', specialWeight='finalweight_Year') ,
    estimate2010 = subpopEstimate(finalData,parameterName, 'Year', '2010','Year 2010', specialWeight='finalweight_Year') ,
    estimate2011 = subpopEstimate(finalData,parameterName, 'Year', '2011','Year 2011', specialWeight='finalweight_Year') ,
    estimate2012 = subpopEstimate(finalData,parameterName, 'Year', '2012','Year 2012', specialWeight='finalweight_Year') ,
    estimate2013 = subpopEstimate(finalData,parameterName, 'Year', '2013','Year 2013', specialWeight='finalweight_Year') ,
    estimate2014 = subpopEstimate(finalData,parameterName, 'Year', '2014','Year 2014', specialWeight='finalweight_Year') ,
    estimate2015 = subpopEstimate(finalData,parameterName, 'Year', '2015','Year 2015', specialWeight='finalweight_Year') ,
    estimate2016 = subpopEstimate(finalData,parameterName, 'Year', '2016','Year 2016', specialWeight='finalweight_Year') ,
    estimate2017 = subpopEstimate(finalData,parameterName, 'Year', '2017','Year 2017', specialWeight='finalweight_Year') ,
    estimate2018 = subpopEstimate(finalData,parameterName, 'Year', '2018','Year 2018', specialWeight='finalweight_Year') ,
    # By Phase
    estimatePhase1 = subpopEstimate(finalData,parameterName, 'Panel', 'Phase1','Phase One 2001-2008', specialWeight='finalweight_Panel1') ,
    estimatePhase2 = subpopEstimate(finalData,parameterName, 'Panel', 'Phase2','Phase Two 2009-2018', specialWeight='finalweight_Panel2') ,
    # Bay/ NonBay
    estimateBay = subpopEstimate(finalData,parameterName, 'BayShed', 'Bay','Bay Watersheds 2001-2018', specialWeight=FALSE),
    estimateNonBay = subpopEstimate(finalData,parameterName, 'BayShed', 'NonBay','Non-Bay Watersheds 2001-2018', specialWeight=FALSE),
    estimateBayPhase1 = subpopEstimate(finalData,parameterName, 'BayPanel', 'BayPhase1','Bay Watersheds 2001-2008', specialWeight='finalweight_Panel1'),
    estimateBayPhase2 = subpopEstimate(finalData,parameterName, 'BayPanel', 'BayPhase2','Bay Watersheds 2009-2018', specialWeight='finalweight_Panel2'),
    estimateNonBayPhase1 = subpopEstimate(finalData,parameterName, 'BayPanel', 'NonBayPhase1','Non-Bay Watersheds 2001-2008', specialWeight='finalweight_Panel1'),
    estimateNonBayPhase2 = subpopEstimate(finalData,parameterName, 'BayPanel', 'NonBayPhase2','Non-Bay Watersheds 2009-2018', specialWeight='finalweight_Panel2'),
    # Sampling Phase
    estimateBioPanelPhase1 = subpopEstimate(finalData,parameterName, 'BioPanel', 'Phase1','VSCI Scores 2001-2004', specialWeight='finalweight_BioPanel1'),
    estimateBioPanelPhase2 = subpopEstimate(finalData,parameterName, 'BioPanel', 'Phase2','VSCI Scores 2005-2008', specialWeight='finalweight_BioPanel2'),
    estimateBioPanelPhase3 = subpopEstimate(finalData,parameterName, 'BioPanel', 'Phase3','VSCI Scores 2009-2013', specialWeight='finalweight_BioPanel3'),
    estimateBioPanelPhase4 = subpopEstimate(finalData,parameterName, 'BioPanel', 'Phase4','VSCI Scores 2014-2018', specialWeight='finalweight_BioPanel4'),
    # IR window
    estimateIR2008 = subpopEstimate(finalData,parameterName, 'IR2008', '2008','IR2008', specialWeight='finalweight_IR2008'),
    estimateIR2010 = subpopEstimate(finalData,parameterName, 'IR2010', '2010','IR2010', specialWeight='finalweight_IR2010'),
    estimateIR2012= subpopEstimate(finalData,parameterName, 'IR2012', '2012','IR2012', specialWeight='finalweight_IR2012'),
    estimateIR2014 = subpopEstimate(finalData,parameterName, 'IR2014', '2014','IR2014', specialWeight='finalweight_IR2014'),
    estimateIR2016 = subpopEstimate(finalData,parameterName, 'IR2016', '2016','IR2016', specialWeight='finalweight_IR2016'),
    estimateIR2018 = subpopEstimate(finalData,parameterName, 'IR2018', '2018','IR2018', specialWeight='finalweight_IR2018'),
    estimateIR2020 = subpopEstimate(finalData,parameterName, 'IR2020', '2020','IR2020', specialWeight='finalweight_IR2020'),

    # Stream size
    estimateSmall = subpopEstimate(finalData,parameterName, 'StreamSizeCat','Small',NA, specialWeight=FALSE),
    estimateMedium = subpopEstimate(finalData,parameterName, 'StreamSizeCat','Medium',NA, specialWeight=FALSE),
    estimateLarge = subpopEstimate(finalData,parameterName, 'StreamSizeCat','Large',NA, specialWeight=FALSE),
    # Stream size and sample phase
    estimatePhase1Small = subpopEstimate(finalData,parameterName, 'StreamSizeCatPhase','Phase1Small',NA, specialWeight='finalweight_Panel1'),
    estimatePhase2Small = subpopEstimate(finalData,parameterName, 'StreamSizeCatPhase','Phase2Small',NA, specialWeight='finalweight_Panel2'),
    estimatePhase1Medium = subpopEstimate(finalData,parameterName, 'StreamSizeCatPhase','Phase1Medium',NA, specialWeight='finalweight_Panel1'),
    estimatePhase2Medium = subpopEstimate(finalData,parameterName, 'StreamSizeCatPhase','Phase2Medium',NA, specialWeight='finalweight_Panel2'),
    estimatePhase1Large = subpopEstimate(finalData,parameterName, 'StreamSizeCatPhase','Phase1Large',NA, specialWeight='finalweight_Panel1'),
    estimatePhase2Large = subpopEstimate(finalData,parameterName, 'StreamSizeCatPhase','Phase2Large',NA, specialWeight='finalweight_Panel2')
  )
}

allCDFresults <- function(designStatus, surveyData, parameter){
  
  # Organize new parameter data
  parameterData <- data.frame(select(surveyData,siteID,parameter))
  parameterData[,2] <- as.numeric(as.character(parameterData[,2])) # change to numeric in case it was saved as anything else
  
  
  # Change status if particular parameter wasnt sampled
  initialWeightData <- left_join(designStatus,parameterData,by='siteID') %>%
    mutate(parameter_status=ifelse(status == "TS" & is.na(!!as.name(parameter)),"OT", # if TS but not sampled, change to OT
                                   ifelse(status == "OT" & is.na(!!as.name(parameter)),"OT", # if OT but not sampled, keep OT
                                          ifelse(status == "PD", "PD", # if PD, keep it
                                                 ifelse(status == 'NT','NT', # if NT, keep it
                                                        ifelse(status == "TS",'TS', NA))))), # if TS AND sampled, keep as TS
           designweightoriginal = as.factor(`strahler order`), # easier to change as factor
           designweightoriginal = dplyr::recode(designweightoriginal,`1`="3790.5165999999999",
                                                `2`="947.62919999999997",
                                                `3`="541.50239999999997",
                                                `4`="315.87639999999999", 
                                                `5`="140.3895", 
                                                `6`="140.3895")) # overwrite all design weights back to original
  
  # Adjust initial weights for trend stations across various data windows
  # First calculate number of times sampled in each window
  
  trendWeightAdjustments <- mutate(initialWeightData,
                                   designweightoriginal = as.numeric(as.character(designweightoriginal)), #factor to numeric
                                   siteIDoriginal=gsub("_.*$", "", siteID)) %>% # get rid of concatenated year for trends to make calculations easier
    # Full window
    group_by(siteIDoriginal) %>% 
    mutate(nYearsSampled = ifelse(is.na(siteIDoriginal),NA,n())) %>%
    ungroup()%>%
    
    # 2020 IR
    group_by(siteIDoriginal, IR2020) %>%
    mutate(nYearsSampled_IR2020 = ifelse(is.na(IR2020),NA,n())) %>%
    ungroup() %>%
    
    # 2018 IR
    group_by(siteIDoriginal, IR2018) %>%
    mutate(nYearsSampled_IR2018 = ifelse(is.na(IR2018),NA,n())) %>%
    ungroup() %>%
    # 2016 IR
    group_by(siteIDoriginal, IR2016) %>%
    mutate(nYearsSampled_IR2016 = ifelse(is.na(IR2016),NA,n())) %>%
    ungroup()%>%
    # 2014 IR
    group_by(siteIDoriginal, IR2014) %>%
    mutate(nYearsSampled_IR2014 = ifelse(is.na(IR2014),NA,n()))%>%
    ungroup()%>%
    # 2012 IR
    group_by(siteIDoriginal, IR2012) %>%
    mutate(nYearsSampled_IR2012 = ifelse(is.na(IR2012),NA,n())) %>%
    ungroup()%>%
    # 2010 IR
    group_by(siteIDoriginal, IR2010) %>%
    mutate(nYearsSampled_IR2010 = ifelse(is.na(IR2010),NA,n())) %>%
    ungroup()%>%
    # 2008 IR
    group_by(siteIDoriginal, IR2008) %>%
    mutate(nYearsSampled_IR2008 = ifelse(is.na(IR2008),NA,n())) %>%
    # Panels
    group_by(siteIDoriginal, Panel1) %>%
    mutate(nYearsSampled_Panel1 = ifelse(is.na(Panel1),NA,n())) %>%
    ungroup()%>%
    group_by(siteIDoriginal, Panel2) %>%
    mutate(nYearsSampled_Panel2 = ifelse(is.na(Panel2),NA,n())) %>%
    ungroup()%>%
    # Biopanels
    group_by(siteIDoriginal, BioPanel1) %>%
    mutate(nYearsSampled_BioPanel1 = ifelse(is.na(BioPanel1),NA,n())) %>%
    ungroup() %>% 
    group_by(siteIDoriginal, BioPanel2) %>%
    mutate(nYearsSampled_BioPanel2 = ifelse(is.na(BioPanel2),NA,n())) %>%
    ungroup() %>% 
    group_by(siteIDoriginal, BioPanel3) %>%
    mutate(nYearsSampled_BioPanel3 = ifelse(is.na(BioPanel3),NA,n())) %>%
    ungroup() %>% 
    group_by(siteIDoriginal, BioPanel4) %>%
    mutate(nYearsSampled_BioPanel4 = ifelse(is.na(BioPanel4),NA,n())) %>%
    ungroup() %>% 
    # Divide out weight by years sampled in each window
    # if the site wasnt sampled in a window, give it the original weight but it will be adjusted according to an OT status later
    mutate(designweight_all= designweightoriginal/nYearsSampled,
           designweight_IR2020= ifelse(is.na(nYearsSampled_IR2020),designweightoriginal,designweightoriginal/nYearsSampled_IR2020),
           designweight_IR2018= ifelse(is.na(nYearsSampled_IR2018),designweightoriginal,designweightoriginal/nYearsSampled_IR2018),
           designweight_IR2016= ifelse(is.na(nYearsSampled_IR2016),designweightoriginal,designweightoriginal/nYearsSampled_IR2016),
           designweight_IR2014= ifelse(is.na(nYearsSampled_IR2014),designweightoriginal,designweightoriginal/nYearsSampled_IR2014),
           designweight_IR2012= ifelse(is.na(nYearsSampled_IR2012),designweightoriginal,designweightoriginal/nYearsSampled_IR2012),
           designweight_IR2010= ifelse(is.na(nYearsSampled_IR2010),designweightoriginal,designweightoriginal/nYearsSampled_IR2010),
           designweight_IR2008= ifelse(is.na(nYearsSampled_IR2008),designweightoriginal,designweightoriginal/nYearsSampled_IR2008),
           designweight_Panel1= ifelse(is.na(nYearsSampled_Panel1),designweightoriginal,designweightoriginal/nYearsSampled_Panel1),
           designweight_Panel2= ifelse(is.na(nYearsSampled_Panel2),designweightoriginal,designweightoriginal/nYearsSampled_Panel2),
           designweight_BioPanel1= ifelse(is.na(nYearsSampled_BioPanel1),designweightoriginal,designweightoriginal/nYearsSampled_BioPanel1),
           designweight_BioPanel2= ifelse(is.na(nYearsSampled_BioPanel2),designweightoriginal,designweightoriginal/nYearsSampled_BioPanel2),
           designweight_BioPanel3= ifelse(is.na(nYearsSampled_BioPanel3),designweightoriginal,designweightoriginal/nYearsSampled_BioPanel3),
           designweight_BioPanel4= ifelse(is.na(nYearsSampled_BioPanel4),designweightoriginal,designweightoriginal/nYearsSampled_BioPanel4))
  
  
  ## Adjust design weights to get final weights
  
  # Initial sample frame inputs
  # List stream order by kilometer it represents
  sframe <- c('1st'=51210, '2nd'=13680, '3rd'=7781.08, '4th'=4448.257, 
              '5th'=1731.302, '6th'=163.901, '7th'=14.7099 )
  
  # recode to factor to make sframe match up to stream order
  trendWeightAdjustments$`strahler order` <- as.factor(trendWeightAdjustments$`strahler order`)
  levels(trendWeightAdjustments$`strahler order`) <- c('1st','2nd','3rd','4th','5th','6th','7th')
  
  finalWeights <- select(trendWeightAdjustments,siteID:IR2020,siteIDoriginal,designweightoriginal,
                         designweight_all:designweight_BioPanel4,
                         parameter_status,!!as.name(parameter))
  finalWeights$finalweight_Year <- adjwgt(rep(TRUE,length(finalWeights$designweightoriginal)),finalWeights$designweightoriginal,
                                          finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_all <- adjwgt(rep(TRUE,length(finalWeights$designweight_all)),finalWeights$designweight_all,
                                         finalWeights$`strahler order`, sframe)
  
  finalWeights$finalweight_IR2020 <- adjwgt(rep(TRUE,length(finalWeights$designweight_IR2020)),finalWeights$designweight_IR2020,
                                            finalWeights$`strahler order`, sframe)
  
  finalWeights$finalweight_IR2018 <- adjwgt(rep(TRUE,length(finalWeights$designweight_IR2018)),finalWeights$designweight_IR2018,
                                            finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_IR2016 <- adjwgt(rep(TRUE,length(finalWeights$designweight_IR2016)),finalWeights$designweight_IR2016,
                                            finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_IR2014 <- adjwgt(rep(TRUE,length(finalWeights$designweight_IR2014)),finalWeights$designweight_IR2014,
                                            finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_IR2012 <- adjwgt(rep(TRUE,length(finalWeights$designweight_IR2012)),finalWeights$designweight_IR2012,
                                            finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_IR2010 <- adjwgt(rep(TRUE,length(finalWeights$designweight_IR2010)),finalWeights$designweight_IR2010,
                                            finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_IR2008 <- adjwgt(rep(TRUE,length(finalWeights$designweight_IR2008)),finalWeights$designweight_IR2008,
                                            finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_Panel1 <- adjwgt(rep(TRUE,length(finalWeights$designweight_Panel1)),finalWeights$designweight_Panel1,
                                            finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_Panel2 <- adjwgt(rep(TRUE,length(finalWeights$designweight_Panel2)),finalWeights$designweight_Panel2,
                                            finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_BioPanel1 <- adjwgt(rep(TRUE,length(finalWeights$designweight_BioPanel1)),finalWeights$designweight_BioPanel1,
                                               finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_BioPanel2 <- adjwgt(rep(TRUE,length(finalWeights$designweight_BioPanel2)),finalWeights$designweight_BioPanel2,
                                               finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_BioPanel3 <- adjwgt(rep(TRUE,length(finalWeights$designweight_BioPanel3)),finalWeights$designweight_BioPanel3,
                                               finalWeights$`strahler order`, sframe)
  finalWeights$finalweight_BioPanel4 <- adjwgt(rep(TRUE,length(finalWeights$designweight_BioPanel4)),finalWeights$designweight_BioPanel4,
                                               finalWeights$`strahler order`, sframe)
  
  
  
  # Change decimal degree coordinates to marinus for equal area projection (needed for spsurvey::cat.analysis())
  marinus <- spsurvey::marinus(finalWeights$`Latitude-DD`,finalWeights$`Longitude-DD`)
  finalWeights <- cbind(finalWeights,data.frame(xmarinus=marinus$x,ymarinus=marinus$y))
  rm(marinus)
  
  ####Stream Extent and Status Estimate
  
  siteExtent <- data.frame(siteID=finalWeights$siteID, Use=rep(TRUE,nrow(finalWeights)) )
  subpopExtent <- data.frame(siteID=finalWeights$siteID,Region=rep('Virginia',nrow(finalWeights)) )
  designExtent <- data.frame(siteID=finalWeights$siteID,stratum=rep(1,nrow(finalWeights)),
                             wgt=finalWeights$finalweight_all, xcoord=finalWeights$xmarinus,ycoord=finalWeights$ymarinus)
  
  StatusTNT <- finalWeights$status
  levels(StatusTNT) <- list(T=c('TS','PD','OT'), NT=c('NT') )
  
  data.cat.ext <- data.frame(finalWeights[,c('siteID','status')],StatusTNT)
  
  popstatus.est <- cat.analysis(sites = siteExtent, subpop = subpopExtent, design = designExtent,
                                data.cat = data.cat.ext, conf=95, vartype="Local")
  
  dataAnalyzed <- filter(finalWeights, parameter_status=='TS') %>%
    select(siteID,!!as.name(parameter),parameter_status,everything())
  
  output <- listOfResults(popstatus.est, dataAnalyzed, parameter)
  
  return(output)
}

```


Below is how one would run a single parameter and unpack the CDF and PCT data into individual data frames. To view the list structure or data, follow script examples.

```{r runIndividualParameter}
LRBS <- allCDFresults(designStatus, surveyData,'LRBS')
LRBS_CDFdf <- suppressWarnings(LRBS[3:length(LRBS)] %>% map_df(1))
LRBS_PCTdf <-  suppressWarnings(LRBS[3:length(LRBS)] %>% map_df(2))


```

Same for VSCI and how to get all CDF estimates at VSCI=60.

```{r VSCIexample}
# VLOOKUP (Excel function hack) by Julin Maloof, edited for list application by Emma Jones
vlookupEVJ <- function(table, #the table where you want to look for it; will look in first column
                       ref, #the value or values that you want to look for
                       column, #the column that you want the return data to come from,
                       range=FALSE, #if there is not an exact match, return the closest?
                       larger=FALSE) #if doing a range lookup, should the smaller or larger key be used?)
{
  # 2020 addition, make tibbles dataframes
  table <- as.data.frame(table)
  
  if(!is.numeric(column) & !column %in% colnames(table)) {
    stop(paste("can't find column",column,"in table"))
  }
  if(range) {
    if(!is.numeric(table[,1])) {
      stop(paste("The first column of table must be numeric when using range lookup"))
    }
    table <- table[order(table[,1]),] 
    index <- findInterval(ref,table[,1])
    if(larger) {
      index <- ifelse(ref %in% table[,1],index,index+1)
    }
    output <- table[index,column]
    output[!index <= dim(table)[1]] <- NA
    
  } else {
    output <- table[match(ref,table[,1]),column]
    output[!ref %in% table[,1]] <- NA #not needed?
  }
  dim(output) <- dim(ref)
  output
}


# run VSCI
VSCI <- allCDFresults(designStatus, surveyData,'VSCIVCPMI')



# find CDF results at 60 for all subpopulations
VSCI60 <- VSCI[3:length(VSCI)] %>% # start with our big 'ol list of lists
  map('CDF') %>% #extract just the 'CDF' list from each list item (subpopulation)
  map(`[`, c('Value','Estimate.P')) %>% # kinda like dplyr::select here, extract the Value and Estimate.P columns from each list item (CDF)
  map( vlookupEVJ, 60  , 2, TRUE) %>% # use the vlookup function on each of the tables to find the Estimate.P where value = 60
  tibble::enframe(value = "CDFestimateAt60") # change the results from that list into a tibble (kinda like dataframe)

View(VSCI60)
```


### Run all parameters

```{r runEverything}
startTime <- Sys.time()
# probably a cooler way to do this with purrr, but here is my one for loop
for(i in 3:length(surveyData)){
  assign(names(surveyData)[i], suppressWarnings(allCDFresults(designStatus, surveyData,names(surveyData)[i])))
   saveRDS(get(names(surveyData)[i]),paste('processedData/RDSarchive/',names(surveyData)[i],'.RDS',sep=''))
}


# Then unpack each parameter to get a long df of CDF and PCT data
# Needs to be in separate loop so names of each object to be called will be in the environment
allCDF <- data.frame(Type=NA,Subpopulation=NA,Indicator=NA,Value=NA,NResp=NA,Estimate.P=NA,StdError.P=NA,LCB95Pct.P=NA,UCB95Pct.P=NA,   
                     Estimate.U=NA,StdError.U=NA,LCB95Pct.U=NA,UCB95Pct.U=NA)
allPCT <- data.frame(Type=NA,Subpopulation=NA,Indicator=NA,Statistic=NA,NResp=NA,Estimate=NA,StdError=NA,LCB95Pct=NA,UCB95Pct=NA)

# okay, one last one, I promise
for(i in 3:length(surveyData)){
  assign(paste(names(surveyData)[i],"CDFdf",sep="_"),
         suppressWarnings(get(names(surveyData)[i])[3:length(get(names(surveyData)[i]))] %>% 
                            map_df(1)))
  assign(paste(names(surveyData)[i],"PCTdf",sep="_"),
         suppressWarnings(get(names(surveyData)[i])[3:length(get(names(surveyData)[i]))] %>%
                            map_df(2)))
  allCDF <- rbind(allCDF,get(paste(names(surveyData)[i],"CDFdf",sep="_")))
  allPCT <- rbind(allPCT,get(paste(names(surveyData)[i],"PCTdf",sep="_")))
  
}

write.csv(allCDF,'processedData/allCDF.csv',row.names = F)
write.csv(allPCT,'processedData/allPCT.csv',row.names = F)

howLong <- Sys.time()-startTime

```




### Relative Risk

Now run relative risk analyses. First, get final weights from VSCIVCPMI all years to use for relative risk analyses.

```{r relativeRiskgetWeights}
weights <- readRDS('processedData/RDSarchive/VSCIVCPMI.RDS') 
weights <- weights[["dataAnalyzed"]] %>% select(siteID, Year, finalweight_all)
```


```{r relativeRiskOrganization}
####Need to remove the fair....need to run code at home to check results
library(magrittr)
toNumeric <- names(select(surveyData,DO:wshdImpPCT))
surveyData[,toNumeric] %<>% lapply(function(x) as.numeric(as.character(x)))
rrisk <- left_join(surveyData,select(designStatus,siteID, Year,`Longitude-DD`,`Latitude-DD`)
                   ,by=c('siteID','Year')) %>%
  select(siteID,Year,`Longitude-DD`,`Latitude-DD`,everything()) %>%
  left_join(weights, by=c('siteID','Year')) %>%
  select(siteID:`Latitude-DD`,finalweight_all,TN,TP,TDS,LRBS,TotHab,VSCIVCPMI,MetalCCU)

rrisk$VSCIstatus <- cut(rrisk$VSCIVCPMI,c(0,50,60,100),labels=c('Poor','Fair','Good'))  
rrisk$TotHabstatus <- cut(rrisk$TotHab, c(0,120,150,200),labels=c('Poor','Fair','Good'))
rrisk$TNstatus <- cut(rrisk$TN, c(0,1,2,100),labels=c('Good','Fair','Poor'))
rrisk$TPstatus <- cut(rrisk$TP, c(0,0.02,0.05,100),labels=c('Good','Fair','Poor'))
rrisk$TDSstatus <- cut(rrisk$TDS, c(0,100,350,20000),labels=c('Good','Fair','Poor'))
rrisk$MetalCCUstatus <- cut(rrisk$MetalCCU, c(0,1,2,100),labels=c('Good','Fair','Poor'))
rrisk$LRBSstatus <- cut(rrisk$LRBS, c(-10,-1,-0.5,0.5,10),labels=c('Poor','Fair','Good','Fair2'))
names(rrisk)
rrisk$VSCIstatus[ rrisk$VSCIstatus == "Fair" ] = NA
rrisk$TotHabstatus[ rrisk$TotHabstatus == "Fair" ] = NA
rrisk$TNstatus[ rrisk$TNstatus == "Fair" ] = NA
rrisk$TPstatus[ rrisk$TPstatus == "Fair" ] = NA
rrisk$TDSstatus[ rrisk$TDSstatus == "Fair" ] = NA
rrisk$MetalCCUstatus[ rrisk$MetalCCUstatus == "Fair" ] = NA
rrisk$LRBSstatus[ rrisk$LRBSstatus == "Fair" ] = NA
rrisk$LRBSstatus[ rrisk$LRBSstatus == "Fair2" ] = NA

```

```{r rriskPrep}

# Create a variable that contains only the name of the response variable.
resp.var<-"VSCIstatus"


#########.
# For stressor variables, we will initially select a few stressor variables 
# All of these must be condition class variables.

stres.vars<- c("TotHabstatus","TNstatus","TPstatus","TDSstatus","MetalCCUstatus","LRBSstatus");

# Create a vector containing the names of all selected stressor variables
#stres.vars<-c("PTL_COND","NTL_COND","TURB_COND","ANC_COND","SALINITY_COND",
#              "RDIS_COND", "RVEG_COND", "LITCVR_COND","LITRIPCVR_COND");


# First, set up the 4 data frames, using columns in ProbMetrics

# 4.1) The sites data frame:
sites.va.rr<-data.frame(siteID=rrisk$siteID, Use=rep(TRUE, nrow(rrisk)))

# ii) The subpopulation data frame. RR and AR estimates have high uncertainties
# and require large sample sizes. Thus, we will estimate RR and AR only for our largest sample size, 
# which is the whole state basis.
subpop.va.rr <- data.frame(siteID=rrisk$siteID, all.virginia=rep("All_of_VA", nrow(rrisk)))


# iii) The design data frame is the same as for extent estimation. However, to be safe,  
#      let's rebuild this data frame, since we are now working from texas.dat.rr;

# add marinus projection coordinates
#tmp <- marinus(rrisk$LatitudeDD,-rrisk$LongitudeDD)
#tmp$xmarinus <- tmp[,'x']
#tmp$ymarinus <- tmp[,'y']
####Noticed USEPA using albers ver marinus projection now (using geodalbers function spsurvey)


albers.cord.rr<-geodalbers(lon=rrisk$`Longitude-DD`,lat=rrisk$`Latitude-DD`)
design.va.rr<-data.frame(siteID=rrisk$siteID, 
                         xcoord=albers.cord.rr$xcoord,ycoord=albers.cord.rr$ycoord, 
                         wgt=rrisk$finalweight_all)


# iv) The data.cat data frame should contain siteID, plus all stressor
# and response variables;
data.cat.va.rr<-subset(rrisk, select=c("siteID",resp.var, stres.vars),drop=T);
names(data.cat.va.rr)[1]<-"siteID";

```

```{r relativeRiskAnalysis}
#Finally, we implement relrisk.analysis().

relrisk.estimates<-relrisk.analysis(sites=sites.va.rr,subpop=subpop.va.rr,
                                    design=design.va.rr,data.rr=as.data.frame(data.cat.va.rr),
                                    response.var=rep(resp.var,length(stres.vars)),
                                    stressor.var=stres.vars,
                                    response.levels=rep(list(c("Poor","Good")),length(stres.vars)), 
                                    stressor.levels=rep(list(c("Poor","Good")),length(stres.vars)))


write.csv(relrisk.estimates, file = "processedData/relriskIR2020.csv", row.names = FALSE)
write.csv(rrisk, file = "processedData/rrisk.csv", row.names = FALSE)

```


### Maps

Now time to make maps of sites sampled across the state for the introduction part of chapter. First you need to make shapefiles of all sites (wadeable and nonwadeable) ever sampled, then split by IR window, then pull watersheds from IR window from Landcover GIS database. I created the GIS project "MapDataOrganization.mxd" to overview the process.

```{r, GISwrangling}

allProb <- left_join(surveyData,designStatus,by=c('siteID','Year')) %>%
  select(siteID, Year, `strahler order`:`Latitude-DD`,Basin:IR2018) %>%
  mutate(StationID = gsub("_.*$", "", siteID)) %>% # make stationID column with no trend years
  select(StationID, everything())%>%
  rename(Longitude=`Longitude-DD`,Latitude=`Latitude-DD`)

# make it a shapefile for maps (sf being a pain with projections right now)
coordinates(allProb) <- ~Longitude+Latitude
proj4string(allProb) <- CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0")

# Now save it
raster::shapefile(allProb, "processedData/allProb2018.shp", overwrite=T)

# Limit to just sites sampled in 2018IR window in GIS
# saved as processedData/IRstations2018.shp

# Bring in watersheds to match to point files
allWshd <- rgdal::readOGR('C:/GIS/ProbMonGIS/DelineatedWatersheds/Watersheds.gdb','AllWatersheds_through2016')
# Get rid of any stray spaces after StationID in attribute table
allWshd@data$StationID <- sub("\r\n" ,"",allWshd@data$StationID) 


allWshdforProb <- allWshd[allWshd@data$StationID %in% allProb@data$StationID,]
allWshdforProb <- spTransform(allWshdforProb,CRS( "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"))

# Now save it
raster::shapefile(allWshdforProb, "processedData/allProbWshd2018.shp", overwrite=T)


# now botable 

# Bring in watershed data
allWshd <- st_as_sf(rgdal::readOGR('C:/GIS/ProbMonGIS/DelineatedWatersheds/Watersheds.gdb','AllWatersheds_through2016')) # bring it in as sf object
# Get rid of any stray spaces after StationID in attribute table
allWshd$StationID <- sub("\r\n" ,"",allWshd$StationID) 

# Limit to just ProbMon wadeable watersheds
allProb_wshd <-  filter(allWshd, StationID %in% allProb$StationID) # allProb_sf$StationID

# Limit to just watersheds sampled in 2018IR window
probIRwshd <- filter(allProb_wshd,StationID %in% probIR_sf$StationID)

# I still need these watersheds in teh GIS database and need to run landcover
#allProb_sf[!(allProb_sf$StationID %in% allWshd$StationID),]$StationID
#probIR_sf[!(probIR_sf$StationID %in% allWshd$StationID),]$StationID

```




Cool, now get boatable sites and watersheds organized (just for map visualization, not for reporting purposes). Working backwards from the designStatus dataframe, take all the NT sites.

```{r boatableSites}
# boatable layers oraganized for last IR
boatable <- rgdal::readOGR('C:/HardDriveBackup/R/IR_2016/data','BoatableSites_20062015')
boatableWshd <- rgdal::readOGR('C:/HardDriveBackup/R/IR_2016/data','IR2016_Bwshds')

boatableSites <- filter(designStatus,status=='NT')%>% # first get all NonTarget
  filter(grepl('Boat',comment) | grepl('Large River',comment) | grepl('Fall only',comment)) %>%
  mutate(StationID = siteID) %>%
  filter(!grepl('VAW',siteID))
boatableSites$StationID <- substr(boatableSites$StationID, start=1, stop=11) # get rid of other characters at end

# now find sites that arent in match boatable existing boatable layer
addToboatable <- unique(boatableSites[!(boatableSites$StationID %in% boatable@data),]$StationID)

boatableSites2018 <- filter(boatableSites,StationID %in% addToboatable)

# Boatable Sites
boatableSites <- boatableSites %>%
  rename(Longitude=`Longitude-DD`,Latitude=`Latitude-DD`)

# make it a shapefile for maps (sf being a pain with projections right now)
coordinates(boatableSites) <- ~Longitude+Latitude
proj4string(boatableSites) <- CRS("+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0")

# Now save it
raster::shapefile(boatableSites, "processedData/boatableSites2018.shp", overwrite=T)

# Now combine "processedData/boatableSites2018.shp" and 'C:/HardDriveBackup/R/IR_2016/data/BoatableSites_20062015.shp' in GIS to get real boatable site layer named 'processedData/boatableSites2006_2016.shp'

boatableSites <- rgdal::readOGR("processedData","boatableSites2006_2016")



# now find watersheds that match boatable StationID's
boatableWshd <- allWshd[allWshd@data$StationID %in% boatableSites$StationID,]
boatableWshd <- spTransform(boatableWshd,CRS( "+proj=longlat +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"))

# Now save it
raster::shapefile(boatableWshd, "processedData/boatableWshd2006_2016.shp", overwrite=T)

```



### Micromap Section

This section makes micromaps from CDF data and shapefiles

```{r}
library(micromap)

# VLOOKUP (Excel function hack) by Julin Maloof
vlookup <- function(ref, #the value or values that you want to look for
                    table, #the table where you want to look for it; will look in first column
                    column, #the column that you want the return data to come from,
                    range=FALSE, #if there is not an exact match, return the closest?
                    larger=FALSE) #if doing a range lookup, should the smaller or larger key be used?)
{
  # 2020 addition, make tibbles dataframes
  table <- as.data.frame(table)
  
  if(!is.numeric(column) & !column %in% colnames(table)) {
    stop(paste("can't find column",column,"in table"))
  }
  if(range) {
    if(!is.numeric(table[,1])) {
      stop(paste("The first column of table must be numeric when using range lookup"))
    }
    table <- table[order(table[,1]),] 
    index <- findInterval(ref,table[,1])
    if(larger) {
      index <- ifelse(ref %in% table[,1],index,index+1)
    }
    output <- table[index,column]
    output[!index <= dim(table)[1]] <- NA
    
  } else {
    output <- table[match(ref,table[,1]),column]
    output[!ref %in% table[,1]] <- NA #not needed?
  }
  dim(output) <- dim(ref)
  output
}

# Stats lookup function (gets data in correct format for micromaps)
statslookup <- function(indicator,measure,category,revOrder){
  chowan <-  filter(dat,Subpopulation=='Chowan'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  rappahannock <-  filter(dat,Subpopulation=='Rappahannock'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  york <-  filter(dat,Subpopulation=='York'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  potomac <-  filter(dat,Subpopulation=='Potomac'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  shenandoah <-  filter(dat,Subpopulation=='Shenandoah'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  roanoke <-  filter(dat,Subpopulation=='Roanoke Basin'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  james <-  filter(dat,Subpopulation=='James Basin'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  new <-  filter(dat,Subpopulation=='New'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  bigsandy <-  filter(dat,Subpopulation=='Big Sandy'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  cp <-  filter(dat,Subpopulation=='Clinch-Powell'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  holston <-  filter(dat,Subpopulation=='Holston'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  Virginia <-  filter(dat,Subpopulation=='Virginia'&Indicator==indicator)%>%
    select(Value,Estimate.P,LCB95Pct.P,UCB95Pct.P,NResp)
  
  x <- data.frame(Subpopulation=c('Chowan','Rappahannock','York','Potomac','Shenandoah','Roanoke','James',
                                  'New','Big Sandy','Clinch-Powell','Holston','Virginia'),
                  Category=category,NResp=c(max(chowan$NResp),max(rappahannock$NResp),max(york$NResp),
                                            max(potomac$NResp),max(shenandoah$NResp),max(roanoke$NResp),
                                            max(james$NResp),max(new$NResp),max(bigsandy$NResp),max(cp$NResp),
                                            max(holston$NResp),max(Virginia$NResp)),
                  matchingdfname=c('chowan','rappahannock','york','potomac','shenandoah','roanoke','james','new','bigsandy','cp','holston','Virginia'))
  y=data.frame(Estimate.P=NA,LCB95Pct.P=NA, UCB95Pct.P=NA)
  for(i in 1:nrow(x)){
    y[i,1] <- vlookup(measure,get(as.character(x[i,4])),2,TRUE)
    y[i,2] <- vlookup(measure,get(as.character(x[i,4])),3,TRUE)
    y[i,3] <- vlookup(measure,get(as.character(x[i,4])),4,TRUE)
    y[is.na(y)] <- 0
    y[y>100] <- 100
  }
  y2 <- mutate(y,Estimate.P2=100-Estimate.P,LCB95Pct.P2=Estimate.P2-(Estimate.P-LCB95Pct.P),
               UCB95Pct.P2=Estimate.P2+(UCB95Pct.P-Estimate.P))%>%select(Estimate.P2,LCB95Pct.P2,UCB95Pct.P2)
  y2[y2>100] <- 100
  names(y) <- c(paste(indicator,"Estimate.P",sep=""),paste(indicator,"LCB95Pct.P",sep=""),paste(indicator,"UCB95Pct.P",sep=""))
  names(y2) <- c(paste(indicator,"Estimate.P",sep=""),paste(indicator,"LCB95Pct.P",sep=""),paste(indicator,"UCB95Pct.P",sep=""))
  if(revOrder==FALSE){z <- cbind(x,y)}else{z<-cbind(x,y2)}
  return(z)
}

```

```{r VSCIstatusMicromap} 
basinssmooth <- readOGR('originalData','VAbasins_smoothNoChesPeeDee') # have to use sp for micromaps
map.table <- create_map_table(basinssmooth,'BASIN')

dat <- read_csv('processedData/allCDF.csv')

# for other VSCI graphic
chowan <-  filter(dat, Subpopulation == 'Chowan' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
rappahannock <-  filter(dat, Subpopulation == 'Rappahannock' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
york <-  filter(dat, Subpopulation == 'York' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
potomac <-  filter(dat, Subpopulation == 'Potomac' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
shenandoah <-  filter(dat, Subpopulation == 'Shenandoah' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
roanoke <-  filter(dat, Subpopulation == 'Roanoke Basin' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
james <-  filter(dat, Subpopulation == 'James Basin' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
new <-  filter(dat, Subpopulation == 'New' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
bigsandy <-  filter(dat, Subpopulation == 'Big Sandy' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
cp <-  filter(dat, Subpopulation == 'Clinch-Powell' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())
holston <-  filter(dat, Subpopulation == 'Holston' & Indicator == 'VSCIVCPMI') %>%
  select(Value:UCB95Pct.P,everything())

# pull median VSCI scores and upper/lower bounds for micromap
stats <- data.frame(Basin=c('Chowan','Rappahannock','York','Potomac','Shenandoah','Roanoke','James','New',
                            'Big Sandy','Clinch-Powell', 'Holston'),Indicator='VSCIVCPMI')

# Switch lookup column to search for median
chowan <- select(chowan,Estimate.P,everything())
rappahannock <- select(rappahannock,Estimate.P,everything())
york <- select(york,Estimate.P,everything())
potomac <- select(potomac,Estimate.P,everything())
shenandoah <- select(shenandoah,Estimate.P,everything())
roanoke <- select(roanoke,Estimate.P,everything())
james <- select(james,Estimate.P,everything())
new <- select(new,Estimate.P,everything())
bigsandy <- select(bigsandy,Estimate.P,everything())
cp <- select(cp,Estimate.P,everything())
holston <- select(holston,Estimate.P,everything())

#statssuperB <- suppressWarnings(
#  rbind(data.frame(vlookup(50,chowan,2:4,TRUE)),data.frame(vlookup(50,rapYork,2:4,TRUE)),
#        data.frame(vlookup(50,potShen,2:4,TRUE)),data.frame(vlookup(50,roanoke,2:4,TRUE)),
#        data.frame(vlookup(50,tennessee,2:4,TRUE)),data.frame(vlookup(50,james,2:4,TRUE)),
#        data.frame(vlookup(50,new,2:4,TRUE))))

statsbasin <- rbind(
  data.frame(x25=vlookup(25,chowan,2,TRUE),x50=vlookup(50,chowan,2,TRUE),
             x75=vlookup(75,chowan,2,TRUE),n=max(chowan$NResp)),
  data.frame(x25=vlookup(25,rappahannock,2,TRUE),x50=vlookup(50,rappahannock,2,TRUE),
             x75=vlookup(75,rappahannock,2,TRUE),n=max(rappahannock$NResp)),
  data.frame(x25=vlookup(25,york,2,TRUE),x50=vlookup(50,york,2,TRUE),
             x75=vlookup(75,york,2,TRUE),n=max(york$NResp)),
  data.frame(x25=vlookup(25,potomac,2,TRUE),x50=vlookup(50,potomac,2,TRUE),
             x75=vlookup(75,potomac,2,TRUE),n=max(potomac$NResp)),
  data.frame(x25=vlookup(25,shenandoah,2,TRUE),x50=vlookup(50,shenandoah,2,TRUE),
             x75=vlookup(75,shenandoah,2,TRUE),n=max(shenandoah$NResp)),
  data.frame(x25=vlookup(25,roanoke,2,TRUE),x50=vlookup(50,roanoke,2,TRUE),
             x75=vlookup(75,roanoke,2,TRUE),n=max(roanoke$NResp)),
  data.frame(x25=vlookup(25,james,2,TRUE),x50=vlookup(50,james,2,TRUE),
             x75=vlookup(75,james,2,TRUE),n=max(james$NResp)),
  data.frame(x25=vlookup(25,new,2,TRUE),x50=vlookup(50,new,2,TRUE),
             x75=vlookup(75,new,2,TRUE),n=max(new$NResp)),
  data.frame(x25=vlookup(25,bigsandy,2,TRUE),x50=vlookup(50,bigsandy,2,TRUE),
             x75=vlookup(75,bigsandy,2,TRUE),n=max(bigsandy$NResp)),
  data.frame(x25=vlookup(25,cp,2,TRUE),x50=vlookup(50,cp,2,TRUE),
             x75=vlookup(75,cp,2,TRUE),n=max(cp$NResp)),
  data.frame(x25=vlookup(25,holston,2,TRUE),x50=vlookup(50,holston,2,TRUE),
             x75=vlookup(75,holston,2,TRUE),n=max(holston$NResp)))

stats <- cbind(stats,statsbasin)

#mmplot(stat.data=stats,
#       map.data=map.table,
#       panel.types=c('dot_legend', 'labels', 'dot_cl', 'map'),
#       panel.data=list(NA,'Basin',list('x50', 'x25', 'x75'),NA),
#       ord.by='x50', grouping=4,
#       median.row=F,
#       map.link=c("Basin", "ID"))



suppressWarnings(
  mmplot(stat.data=stats,
       map.data=map.table,
       map.link=c("Basin", "ID"),
       panel.types=c('dot_legend', 'labels','labels', 'dot_cl', 'map'),
       panel.data=list(NA,'Basin','n',list('x50', 'x25', 'x75'),NA),
       ord.by='x50',
       grouping=4,
       median.row=F,
       plot.height=7,
       plot.width=8,
       colors=brewer.pal(7, "Spectral"),
       rev.ord=T,
       panel.att=list(list(1, point.type=20, point.border=TRUE, point.size=2),
                      list(2, header='Basin', panel.width=.4, 
                           align='left', text.size=.9),
                      list(3,header='n',panel.width=.2,align='left',text.size=.9),
                      list(4, header='Estimated Median VSCI Score and \nAssociated Interquartile Range',
                           graph.bgcolor='lightgray', point.size=1.5,
                           xaxis.ticks=list(40,50,60,70,80), xaxis.labels=list(40,50,60,70,80)
                           ,add.line=60,add.line.col='black',add.line.typ='dashed',
                           xaxis.title='VSCI Score'),
                      list(5, header='Light Gray Means\nPreviously Displayed',
                           map.all=TRUE, fill.regions='aggregate',
                           active.border.color='black', active.border.size=1.0,
                           inactive.border.color=gray(.7), inactive.border.size=1, 
                           panel.width=1.0))) )

 


```




```{r}
TotHabstats <- statslookup('TotHab',120,'SubOptimal',FALSE)%>%select(-c(matchingdfname))

VSCIstats <- statslookup('VSCIVCPMI',60,'SubOptimal',FALSE)%>%select(-c(matchingdfname))


TotHaball <- merge(VSCIstats,TotHabstats,by=c('Subpopulation','Category'))
TotHabsummary <- TotHaball%>%
  filter(Subpopulation!='Virginia')%>% # get rid of Virginia to make mmplot work
  arrange(desc(TotHabEstimate.P))

map.table <- create_map_table(basinssmooth,'BASIN')
suppressWarnings(
  mmplot(stat.data=TotHabsummary,
       map.data=map.table,
       map.link=c("Subpopulation", "ID"),
       panel.types=c('map','labels', 'bar_cl', 'bar_cl'),
       panel.data=list(NA,'Subpopulation',
                       list( "TotHabEstimate.P","TotHabLCB95Pct.P","TotHabUCB95Pct.P"),
                       list("VSCIVCPMIEstimate.P","VSCIVCPMILCB95Pct.P","VSCIVCPMIUCB95Pct.P")),
       ord.by='TotHabEstimate.P',
       grouping=3,
       median.row=F,
       plot.height=7,
       plot.width=8,
       colors=brewer.pal(3, "Spectral"),
       rev.ord=T,
       panel.att=list(list(1,header='Light Gray Means\nPreviously Displayed',
                           map.all=TRUE, fill.regions='aggregate',
                           active.border.color='black', active.border.size=1.0,
                           inactive.border.color=gray(.7), inactive.border.size=1, 
                           panel.width=1.0),
                      list(2, header='Basin', panel.width=.4, 
                           align='left', text.size=.9),
                      list(3,header='Percent of Stream Miles with \nSuboptimal Habitat Disturbance',
                           graph.bgcolor='lightgray',
                           graph.bar.size = .4,
                           xaxis.ticks=list(0,20,40,60,80,100), xaxis.labels=list(0,20,40,60,80,100)
                           ,add.line=16.7393531 ,add.line.col='black',add.line.typ='dashed',
                           xaxis.title='Percent of Stream Miles'),
                      list(4,header='Percent of Stream Miles below \n VSCI/VCPMI Assessment Threshold',
                           graph.bgcolor='lightgray',
                           graph.bar.size = .4,
                           add.line=45.41627,add.line.col='black',add.line.typ='dashed',
                           xaxis.title='Percent of Stream Miles',
                           xaxis.ticks = c(0,20,40,60,80,100),
                           xaxis.labels = c(0,20,40,60,80,100)))))


```




